---
title: "Can filled pauses tell you alot about someone?"
author: "Josef Fruehwald"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Can filled pauses tell you alot about someone?}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

> tl;dr: Despite the relatively strong effect of gender on filled pause preference, filled pauses are not a very informative signal of your gender.

So, there's been a lot media attention surrounding a series of Language Log posts on filled pauses that I have also contributed to.
Mark Liberman [has already posted](http://languagelog.ldc.upenn.edu/nll/?p=15048) about how some of the coverage has a sort of "whisper down the lane" quality.
It all bears a resemblance to things we've written and said, but with plenty of random perturbation.

What really interested me about some of the headlines, though, was the recurring theme that your choice of filled pause could "reveal" something about you as a speaker. The Scotsman headline, for instance, read

> [It’s…er…really revealing…y’know](http://www.scotsman.com/news/lori-anderson-it-s-er-really-revealing-y-know-1-3569856)

It gets a bit more extreme than that just under the headline.

> Lori Anderson finds meaningless sounds we make to fill pauses could determine if we’ll vote Ukip or pen a No 1 hit

This is like some kind of bizarro world Sapir-Whorf. It's like correlation, causality and prediction can run in any which direction! Here's the best thumbnail description of what we all *have* found.

- It looks like gender and age are predictive of the frequency with which someone will use "um" or "uh" when they use a filled pause.

That isn't the same thing as saying that your filled pause usage predicts your gender and age, and gender and age are only some predictors for voting patterns, and No 1 hitmaking ability!
And let's just real quick think about what it is your filled pause use is supposedly revealing: just gender and age, perhaps two of the most aggressively signaled social statuses there are already!
It's hard to see what unique information your filled pause preference could add to the signal.

But, I thought it would be interesting to try to calculate how good of a signal your filled pause preference is for something like gender.
There are ways to calculate how much information there is to communicate in a message, and how much information a particular signal can carry.
These estimations involve calculating probabilities over possible messages, and that's where I need to provide a really big caveat that the rest of this analysis is fairly broken.
To put it in STEMy terms, there is more than 1 bit of entropy to gender expression.
But all I have to work with is whether field workers over the past 40 years jotted down an "m" or an "f" on interview report forms.
But just like any other kind of analysis based on imperfect data, the results here are probably going to approximate the situation in reality.

```{r message = F}
  library("UhUm")
  load_install("dplyr")
  load_install("magrittr")
  load_install("reshape2")
  load_install("ggplot2")
  load_install("mgcv")
  load_install("knitr")
  load_install("babynames")

  opts_chunk$set(message = F)
```


First things first.
Here's the PNC data, where I've calculated the average rate of all filled pauses per 1000 words for each decade of date of birth we have.
The point of doing this is to show that the over-all rate of filled pause use is pretty noisy, maybe declining a little bit with more recent generations, and probably no real relationship to gender.

```{r fig.width = 8/1.25, fig.height = 5/1.25}
  um_PNC%>%
    filter(word %in% c("UH","UM"))%>%
    mutate(DOB = year-age,
           decade = floor(DOB/10)*10) -> um_uh

  um_uh%>%
    group_by(decade, sex, idstring)%>%
    summarise(n = n(),
              nwords = max(nwords))%>%
    summarise(fp_freq = exp(mean(log(n/nwords)))*1000)%>%
    filter(decade > 1880)%>%
    ggplot(., aes(decade, fp_freq, color = sex))+
      geom_point()+
      geom_line()+
      ylab("filled pause frequency per 1000 words")+
      scale_color_brewer("gender", palette = "Dark2")+
      expand_limits(y=0)+
      theme_bw()    

```


```{r}

  entropy <- function(probs){
    sum(probs * log2(1/probs), na.rm = T)
  }


  um_uh%>%
    mutate(is_um = (word == "UM")*1,
           dob= year-age,
           decade = floor((year-age)/10)*10,
           schooling = as.numeric(as.character(schooling)))%>%
    group_by(decade,dob,sex, idstring, schooling)%>%
    summarise(is_um = mean(is_um))->speaker_means

  speaker_means %>%
    group_by(decade, sex)%>%
    summarise(is_um = mean(is_um), 
              n = n())%>%
    mutate(n = sum(n))%>%
    filter(n > 5) -> um_probs

  um_probs %>%
    summarise(joint = entropy(c(is_um * 0.5, (1-is_um)*0.5)),
              sex = entropy(c(0.5, 0.5)),
              um = entropy(c(sum(is_um * 0.5), sum((1-is_um)*0.5))),
              um_prob = sum(is_um * 0.5),
              mutual_info = sex + um - joint)->entropies
  

```

Now, here's what's been the topic of conversation. 
This is the probability of using "Um" or "Uh" when you do use a filled pause as a function of your date of birth and gender.
We're still investigating what's gointg on here, but I have to say this graph looks like a healthy change in progress.
It's got a nice S-shape, and women are somewhere between 20 and 30 years in advance of men.

```{r fig.width = 8/1.25, fig.height = 5/1.25}
  ggplot(um_probs, aes(decade+5, is_um, color = sex))+
    geom_point(data = speaker_means, aes(x = dob), size = 1)+
    geom_point()+
    geom_line()+
    theme_bw()+
    geom_hline(y = c(0,1), color = "grey40")+
    xlab("date of birth")+
    scale_color_brewer(palette = "Dark2")+
    theme_bw()
```

The tempting thing to do next, which a lot of the media reports have done, is to invert this figure.
The reasoning goes that since gender and age are good predictors of filled pause preference, your filled pause preference is good signal of your gender and age.
But it doesn't really work like that.

Lets focus on this idea of your filled pauses "revealing" aspects of your identity.
This seems to be explicitly framing the problem in terms of signalling, so I'll be analyzing the data in terms of its **entropy** and **mutual information**.
To simplify things a bit,

- The entropy of a message is a measure of how much information there is to communicate.
- The mutual information between a signal and a message is a measure of how much information the signal is successfully carrying.

Without getting into details, the entropy of a message depends on the probability 


```{r fig.width = 10/1.25, fig.height = 5/1.25}
  ggplot(entropies, aes(decade)) + 
    geom_ribbon(aes(ymax = 1, ymin = 0), alpha = 0.2, color = "grey30") + 
    geom_line(aes(y = mutual_info, color = "mutual\ninformation"), size = 1)+
    geom_ribbon(aes(min = 0, ymax = mutual_info, fill = "mutual\ninformation"), color = "black")+
    ylab("bits")+
    theme_bw()+
    scale_color_brewer(name = "", palette = "Dark2")+
    scale_fill_brewer(name = "", palette = "Dark2")+
    ggtitle("marginal entropy - gender")
```

<!--
```{r fig.width = 10/1.25, fig.height = 5/1.25}
  ggplot(entropies, aes(decade)) + 
    geom_ribbon(aes(ymax = um, ymin = 0), alpha = 0.3, color = "grey30") + 
    geom_line(aes(y = mutual_info, color = "mutual\ninformation"), size = 1)+
    ylab("bits")+
    scale_color_brewer(name = "", palette = "Dark2")+
    theme_bw()+
    ggtitle("marginal entropy - filled pause")
```
-->

```{r}
babynames%>%
  mutate(last_letter = gsub("[A-Za-z]+([a-z]$)", "\\1", name),
         first_letter = gsub("(^[A-Z]).*", "\\1", name),
         decade = floor(year/10)*10) -> baby2
   

baby2%>%
  group_by(decade,sex, name)%>%
  summarise(n2 = sum(n))%>%
  group_by(decade)%>%
  summarise(joint_entropy= entropy(n2/sum(n2))) -> name_joint_entropy

baby2%>%
  group_by(decade, name)%>%
  summarise(n2 = sum(n))%>%
  summarise(margin_letter = entropy(n2/sum(n2)))->name_margin_entropy

name_entropies <- merge(name_joint_entropy, name_margin_entropy)%>%
                    mutate(signal = "name",
                           mutual_info = margin_letter + 1 - joint_entropy)
```

```{r}
baby2%>%
  group_by(decade,sex, first_letter)%>%
  summarise(n2 = sum(n))%>%
  group_by(decade)%>%
  summarise(joint_entropy= entropy(n2/sum(n2))) -> firstL_joint_entropy

baby2%>%
  group_by(decade, first_letter)%>%
  summarise(n2 = sum(n))%>%
  summarise(margin_letter = entropy(n2/sum(n2)))->firstL_margin_entropy

firstL_entropies <- merge(firstL_joint_entropy, firstL_margin_entropy)%>%
                    mutate(signal = "first_letter",
                           mutual_info = margin_letter + 1 - joint_entropy)
```

```{r}
baby2%>%
  group_by(decade,sex, last_letter)%>%
  summarise(n2 = sum(n))%>%
  group_by(decade)%>%
  summarise(joint_entropy= entropy(n2/sum(n2))) -> lastL_joint_entropy

baby2%>%
  group_by(decade, last_letter)%>%
  summarise(n2 = sum(n))%>%
  summarise(margin_letter = entropy(n2/sum(n2)))->lastL_margin_entropy

lastL_entropies <- merge(lastL_joint_entropy, lastL_margin_entropy)%>%
                    mutate(signal = "last_letter",
                           mutual_info = margin_letter + 1 - joint_entropy)
```

```{r}
  entropies <- entropies %>% mutate(signal = "filled_pause")  
  
  all_signals <- rbind_list(name_entropies, firstL_entropies, lastL_entropies, entropies)

```


```{r fig.width = 10/1.25, fig.height = 5/1.25}
ggplot(all_signals, aes(decade, mutual_info)) + 
  geom_ribbon(aes(ymax = 1, ymin = 0), alpha = 0.3, color = "grey30") +   
  #geom_line(size = 1) + 
  geom_ribbon(data = all_signals %>% filter(signal == "name"), color = "black", 
              aes(ymin = 0, ymax = mutual_info, fill = signal))+
  geom_ribbon(data = all_signals %>% filter(signal == "last_letter"), color = "black", 
              aes(ymin = 0, ymax = mutual_info, fill = signal))+  
  geom_ribbon(data = all_signals %>% filter(signal == "first_letter"), color = "black", 
              aes(ymin = 0, ymax = mutual_info, fill = signal))+  
  geom_ribbon(data = all_signals %>% filter(signal == "filled_pause"), color = "black", 
              aes(ymin = 0, ymax = mutual_info, fill = signal))+  
  ylim(0,1) + 
  scale_color_brewer(palette = "Dark2")+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

```


```{r fig.width = 10/1.25, fig.height = 5/1.25}
  um_probs %>%
    dcast(decade ~ sex, value.var = "is_um")%>%
    mutate(um = (f * 0.5)/((f * 0.5)+(m*0.5)),
           uh = ((1-f)*0.5)/(((1-f)*0.5) + ((1-m)*0.5)))%>%
    melt(id = 1, measure.vars = c("um", "uh"))%>%
    ggplot(., aes(decade, value, color = variable))+
      geom_hline(y = c(0, 1))+
      geom_hline(y = 0.5, color = "grey30", linetype = 2)+
      geom_point()+
      geom_line()+
      ylim(0,1)+
      theme_bw()+
      scale_color_brewer(palette = "Set1")
```

```{r fig.width = 10/1.25, fig.height = 5/1.25}
  um_probs %>%
    dcast(decade ~ sex, value.var = "is_um")%>%
    mutate(um = (f * 0.5)/((f * 0.5)+(m*0.5)),
           uh = ((1-f)*0.5)/(((1-f)*0.5) + ((1-m)*0.5)))%>%
    melt(id = 1, measure.vars = c("um", "uh"))%>%
    filter(variable == "um")%>%
    left_join(um_probs)->um_update

  ggplot(um_update, aes(is_um, value, color = sex))+
    geom_point()+
    geom_path()+
    xlim(0,1)+
    ylim(0.5,1)+
    coord_fixed()
```


<!--
```{r}
  early <- um_uh%>%mutate(dob = year-age) %>% filter(dob >= 1900, dob < 1910)
  late <- um_uh %>%mutate(dob = year-age) %>% filter(dob >= 1980, dob < 1990)

  um_probs %>%
    mutate(fp = "um")%>%
    dcast(decade ~ sex+fp, value.var = "is_um")%>%
    mutate(f_uh = 1-f_um,
           m_uh = 1-m_um)->probs

  um_probs %>%
    filter(decade == 1980)%>%
    mutate(fp = "um")%>%
    dcast(decade ~ sex+fp, value.var = "is_um")%>%
    mutate(f_uh = 1-f_um,
           m_uh = 1-m_um)->late_probs


  update_once <- function(word, prior, probs){
    out <- 0
    f_um = probs["f_um"]
    m_um = probs["m_um"]
    f_uh = probs["f_uh"]
    m_uh = probs["m_uh"]
    if(word == "UH"){
      out <- (f_uh * prior) / ((f_uh * prior) + (m_uh * (1-prior)))
    }else if(word == "UM"){
      out <- (f_um * prior) / ((f_um * prior) + (m_um * (1-prior)))      
    }
    return(out)
  }
  update_vector <- function(words, probs){
    out_vector <- rep(0, length = length(words))
    out_vector[1] <- update_once(words[1], 0.5, probs)
    for(i in 2:length(words)){
      out_vector[i] <- update_once(words[i], out_vector[i-1], probs)
    }
    return(out_vector)
  }

  um_uh%>%
    group_by(idstring, sex)%>%
    mutate(dob = year-age,
           decade = floor(dob/10)*10)%>%
    filter(n() > 1)%>%
    left_join(probs)%>%
    mutate(prob_f = update_vector(words = word, probs = c(f_uh = f_uh[1], f_um = f_um[1],
                                                  m_uh = m_uh[1], m_um = m_um[1])))->prob_f
  
```

```{r fig.width = 10/1.25, fig.height = 5/1.25}
  prob_f %>%
    group_by(idstring, sex, dob)%>%
    filter(start_time == max(start_time))%>%
    mutate(cat_f = (prob_f > 0.5)*1)%>%
    ggplot(., aes(dob, cat_f, color = sex))+
      geom_point(aes(y = prob_f))+
      stat_smooth(method = glm, family = binomial)+
      scale_color_brewer(palette = "Dark2")+
      theme_bw()


```

-->