---
title: "Contributing"
author: "Josef Fruehwald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

If you want to contribute data to the UhUm package, first, fork the repository on Github and clone it. Secondly, I would recommend installing the `devtools` and `roxygen2` packages, which are what I have used to create this package in the first place.

## Documenting Data Generation

I would recommend putting any scripts that were used in generating data in an appropriate directory in `inst`, and saving the raw tab delimited data to `inst/extdata`.

## Adding the data to the package

To add the data to the package, first load the data into R and assign it to a descriptively named object. Then, save that object to `data` as an rda file with the same name. For the PNC data, that looked like this:

```{r eval = F}
um_PNC <- read.delim("inst/extdata/PNC_uh_um_demographics.txt")
save(um_PNC, "data/um_PNC.rda")
```

Then, add documentation for the data set to the `R/data.R` file. The PNC documentation looks like this:

```{r eval = F}
#' Uh and Um data from the PNC
#' @format a data frame with 19 columns. Speech chunks were defined by 
#' continuous speech bordered on each side by a pause of 200 miliseconds or more
#'  \describe{
#'    \item{idstring}{Unique id string for each speaker}
#'    \item{word}{\code{UM} or \code{UH}}
#'    \item{start_time}{Onset time of filled pause, in seconds}
#'    \item{end_time}{Offset time of filled pause, in seconds}
#'    \item{vowel_start}{Onset time of filled pause vowel}
#'    \item{vowel_end}{Offset time of filled pause vowel}
#'    \item{nasal_start}{Onset time of nasal, if \code{UM}}
#'    \item{nasal_end}{Offset time of nasal, if \code{UM}}
#'    \item{chunk_start}{Onset time of speech chunk}
#'    \item{chunk_end}{Offset time of speech chunk}
#'    \item{nwords}{Total number of words from this speaker}
#'    \item{sex}{Sex of speaker}
#'    \item{year}{Year of interview}
#'    \item{age}{Age of speaker at time of interview}
#'    \item{ethnicity}{A number of single character codes for speakers' ethnicity}
#'    \item{schooling}{Number of years of schooling}
#'    \item{transcribed}{Total number of transcribed seconds of interview}
#'    \item{total}{Total length of recording}
#'    \item{nvowels}{Number of measured stressed vowels (irrelevant to UM/UH)}
#'  }
#'  
#' @source Philadelphia Neighborhood Corpus
"um_PNC"
```

Just add a few new lines, and enter your own documentation. It must include the name of the object in quotes following the documentation block.

For more information on documenting data, see here: [http://r-pkgs.had.co.nz/data.html](http://r-pkgs.had.co.nz/data.html)

## Document your contribution

In the `Description` file, add your information in the `Authors@R` vector. It should look something like this:

    Authors@R: c(
              person("Josef", "Fruehwald", email = "josef.frueh@ed.ac.uk",
                      role = c("aut", "cre")),
              person("John", "doe", email = "fakeperson@university.edu".
                      role = "aut")
            )

## Document your data

If you are using devtools, use the function `use_vignette()` to generate an R markdown vignette in the `vignettes` directory.