---
title: "The PNC Data"
author: "Josef Fruehwald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The PNC Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

# The raw data

If you've installed the R package, the raw PNC data with demographic information can be found at `system.file("extdata/PNC_uh_um_demographics.txt", package = "UhUm")`. You should also be able to access it directly in an R session.

Packages used in this vignette:
```{r message=FALSE}
  library(UhUm)
  load_install("dplyr")
  load_install("reshape2")
  load_install("ggplot2")
  load_install("mgcv")
  load_install("knitr")
  load_install("lme4")
  opts_chunk$set(message = F)  
```

A glimpse at the data:
```{r}
  um_PNC %>% glimpse()
```


## Um Preference

First, the basic effect, looking just at "um" preference. It's a pretty wopping date of birth effect.

```{r}
  um_uh <- um_PNC%>%
              filter(word %in% c("UM","UH"))%>%
              mutate(DOB = year-age,
                     is_um = (word == "UM")*1)
```

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%"}
  um_uh %>%
    mutate(DOB = year-age,
           dob_decade = floor(DOB/10)*10,
           age_decade = floor(age/10)*10,
           # this coerces a logical to numeric
           is_um = (word == "UM")*1)%>%
    group_by(age, year, DOB, sex, age_decade, dob_decade, idstring)%>%
    summarise(um_pref = mean(is_um),
              n = n())->speaker_prefs

  ggplot(speaker_prefs, aes(DOB, um_pref, color = sex))+
    geom_point(aes(size = n))+
    stat_smooth(se = F, aes(weight = log(n)))+
    scale_size_area()+
    theme_bw()
```

Trying to tease apart generational effects from lifespan effect. Each date of birth cohort looks fairly flat. 

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%"}
  speaker_prefs %>%
    group_by(sex, age_decade, dob_decade)%>%
    summarise(um_pref = mean(um_pref),
              n = n())%>%
    ggplot(., aes(age_decade, um_pref, color = sex))+
      geom_point(aes(size = n))+
      geom_line(aes(group = sex))+
      scale_size_area()+
      facet_wrap(~dob_decade)

```


Estimating the gender effect in each date of birth cohort. Looks to be narrowing as people converge on maximal "UM".

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%"}
  speaker_prefs%>%
    group_by(sex, dob_decade)%>%
    summarise(um_pref = mean(um_pref))%>%
    dcast(dob_decade ~ sex)%>%
    mutate(ratio = f/m)%>%
      ggplot(., aes(dob_decade, f/m))+
        geom_point()+
        geom_line()+
        geom_hline(y = 1)+
        scale_y_continuous(trans = "log2")+
        theme_bw()
```


Just messing around to see if "uh" to "um" could be construed as a gradual or gradient adding on of the final nasal. Doesn't really look like it.
```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%"}
  um_uh%>%
    filter(word == "UM")%>%
    mutate(DOB = year-age,
           nasal_ratio = (nasal_end-nasal_start)/(vowel_end-vowel_start))%>%
    group_by(DOB, sex, idstring)%>%
    summarise(log_mean = mean(log(nasal_ratio)))%>%
    ggplot(., aes(DOB, log_mean, color = sex))+
      geom_point()+
      stat_smooth(se = F)+
      theme_bw()
```


## Location in discourse

So how about this turn inital "um"? The PNC isn't annotated for turns exactly, but looking just at whether it's speech chunk initial, it looks like there *is* an effect, but it's not gender stratified, and affects both words.

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  initial_um <- um_uh%>%
                    #filter out co-extensive ums
                    filter(!(start_time == chunk_start & end_time == chunk_end))%>%
                    mutate(initial = (chunk_start == start_time)*1)

  initial_um %>%
      group_by(idstring, word, age, year, sex) %>%
      summarise(prop_initial = mean(initial))%>%
      ggplot(., aes(year-age, prop_initial, color = sex, shape = word, linetype = word))+
        geom_point()+
        theme_bw()+
        ggtitle("Raw Data")

  initial_um %>%
      mutate(dob_decade = (floor((year-age)/10))*10)%>%
      group_by(word, dob_decade, sex, idstring) %>%
      summarise(prop_initial = mean(initial))%>%
      summarise(prop_initial = mean(prop_initial))%>%
      ggplot(., aes(dob_decade, prop_initial, color=sex, linetype = word, shape = word))+
        geom_point()+
        geom_line()+
        theme_bw()+
        ggtitle("By Decade Averages")
```

If we exclude absolute initial cases, are filled pauses drifting towards the front of speech chunks? Maybe? The slope is kind of weak though.

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh %>%
    filter(chunk_start != start_time)%>%
    mutate(chunk_dur = chunk_end - chunk_start,
           rel_start = start_time - chunk_start,
           prop_start = rel_start/chunk_dur)%>%
    group_by(age, year, sex, word, idstring)%>%
    summarise(prop_start = mean(prop_start))%>%
    ggplot(., aes(year-age, prop_start, color = sex, shape = word, linetype = word))+
      geom_point()+
      geom_hline(y = c(0,1))+
      stat_smooth(se = F)+
      theme_bw()
```

It looks a bit like "Um" comes a little later in a chunk than "Uh".

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh %>%
    filter(chunk_start != start_time)%>%
    mutate(chunk_dur = chunk_end - chunk_start,
           rel_start = start_time - chunk_start,
           prop_start = rel_start/chunk_dur)%>%
    group_by(age, year, sex, word, idstring)%>%
    summarise(prop_start = mean(prop_start))%>%
    dcast(idstring + year + age + sex ~ word, value.var = "prop_start")%>%
    ggplot(., aes(sex, UM-UH, color = sex))+
      geom_point(position = position_jitter(width = 0.1, height = 0), alpha = 0.6)+
      ## Mean and bootstrapped error
      stat_summary(fun.data = mean_cl_boot,
                   B = 10000,
                   size = 1,
                   color = "black")+
      geom_hline(y = 0)+
      theme_bw()
```

## Duration of Following Pauses

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh %>%
        mutate(DOB = year-age,
           dob_decade = floor(DOB/10)*10,
           age_decade = floor(age/10)*10,
           # this coerces a logical to numeric
           fol_pause = (next_seg == "sp")*1)%>%
    group_by(age, year, DOB, sex, age_decade, word, dob_decade, idstring)%>%
    summarise(fol_pause = mean(fol_pause))%>%
    ggplot(., aes(DOB, fol_pause, color = sex, linetype = word))+
      geom_point()+
      stat_smooth(se = F)+
      facet_wrap(~word)
              
```


```{r fig.width = 12/1.25, fig.height = 5/1.25, out.width = "100%", fig.show = 'hold'}
  um_uh$Fol_code <- "misc"
  um_uh$Fol_code[grepl("[A-Z]", um_uh$next_seg)] <- "consonant"
  um_uh$Fol_code[grepl("[AEIOU]", um_uh$next_seg)] <- "vowel"
  um_uh$Fol_code[um_uh$next_seg  == "sp"] <- "pause"

  um_uh %>%
            mutate(DOB = year-age,
           dob_decade = floor(DOB/10)*10,
           age_decade = floor(age/10)*10,
           # this coerces a logical to numeric
           fol_pause = (next_seg == "sp")*1,
           is_um = (word == "UM") * 1)%>%
      filter(Fol_code != "misc" )%>%
      group_by(age, year, DOB, sex, age_decade, Fol_code, dob_decade, idstring)%>%
      summarise(is_um = mean(is_um)) -> speaker_pref_context

  speaker_pref_context%>%
    group_by(sex, dob_decade, Fol_code)%>%
    summarise(is_um = mean(is_um)) -> decade_mean_context

  ggplot(speaker_pref_context, aes(DOB, is_um, color = Fol_code)) + 
      geom_point()+
      geom_point(data = decade_mean_context, aes(dob_decade), size = 3)+
      geom_line(data = decade_mean_context, aes(dob_decade))+
      facet_wrap(~sex)+
      scale_color_brewer(palette = "Dark2")+
      theme_bw()
```


```{r fig.width = 12/1.25, fig.height = 5/1.25, out.width = "100%", fig.show = 'hold'}

  ggplot(speaker_pref_context, aes(DOB, is_um, color = sex)) + 
      geom_point()+
      geom_point(data = decade_mean_context, aes(dob_decade), size = 3)+
      geom_line(data = decade_mean_context, aes(dob_decade))+
      facet_wrap(~Fol_code)+
      scale_color_brewer(palette = "Dark2")+
      theme_bw()
```

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh %>%
     mutate(DOB = year-age,
            DOB0 = DOB - 1950,
           dob_decade = floor(DOB/10)*10,
           age_decade = floor(age/10)*10,
           # this coerces a logical to numeric
           fol_pause = (next_seg == "sp")*1,
           is_um = (word == "UM") * 1)%>%
      filter(sex == "f")->um_to_model

  mod <- glmer(is_um ~ DOB0 * Fol_code + (1| idstring),
               family = binomial, 
               data = um_to_model)#,
               #control=glmerControl(optimizer = "bobyqa"))

  summary(mod)
```


```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh%>%
    filter(next_seg  == "sp")%>%
    mutate(DOB= year-age,
           seg_dur = next_seg_end - next_seg_start)%>%
    group_by(age, year, DOB, sex, word, idstring)%>%
    summarise(seg_dur = log10(mean(seg_dur)))%>%
    ggplot(., aes(DOB, seg_dur, color = word))+
      stat_smooth()
```

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}

  um_uh%>%
    filter(next_seg  == "sp")%>%
    mutate(DOB= year-age,
           seg_dur = next_seg_end - next_seg_start)%>%
    group_by(age, year, DOB, sex, word, idstring)%>%
    summarise(seg_dur = log10(mean(seg_dur)))%>%
    ggplot(., aes(word, seg_dur))+
      geom_jitter()+
      stat_summary(fun.data = mean_cl_boot, color = "red")  


```

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh%>%
    filter(next_seg  == "sp")%>%
    mutate(DOB= year-age,
           seg_dur = next_seg_end - next_seg_start)%>%
    group_by(age, year, DOB, sex, word, idstring)%>%
    summarise(seg_dur = log10(mean(seg_dur)))%>%
    ggplot(., aes(word, seg_dur, color = sex))+
      stat_summary(fun.data = mean_cl_boot)  
```

```{r}
  um_uh%>%
    filter(next_seg  == "sp")%>%
    mutate(DOB= year-age,
           decade = floor(DOB/10)*10,
           seg_dur = next_seg_end - next_seg_start)%>%
    group_by(age, decade, year, DOB, sex, word, idstring)%>%
    summarise(seg_dur = log10(mean(seg_dur)))%>%
    group_by(sex,decade, word)%>%
    summarise(seg_dur = mean(seg_dur))%>%
    summarise(diff = diff(seg_dur))%>%
    ggplot(., aes(decade, diff, color = sex))+
      geom_point()+
      geom_line()
```


```{r}
  um_uh%>%
    filter(next_seg  == "sp")%>%
    mutate(DOB= year-age,
            DOB0 = DOB - 1950,
           seg_dur = next_seg_end - next_seg_start,
           seg_dur0 = log2(seg_dur)-median(log2(seg_dur)),
           is_um = (word == "UM")*1)%>%
    filter(seg_dur < 3)->dur_effect_data

  dur_mod <- glmer(is_um ~ seg_dur0*DOB0*sex + (seg_dur0|idstring), data = dur_effect_data, family = binomial,
                   glmerControl(optimizer="bobyqa"))
  dur_mod2 <- glmer(is_um ~ seg_dur0*sex + DOB0*sex + (seg_dur0|idstring), data = dur_effect_data, family = binomial,
                    glmerControl(optimizer="bobyqa"))


  anova(dur_mod, dur_mod2)
```

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  dur_effect_data %>%
    mutate(decade = floor(DOB/10)*10,
           seg_dur_cat = cut(seg_dur0, 5))%>%
    group_by(decade, seg_dur_cat, sex, idstring)%>%
    summarise(is_um = mean(is_um), n = n())%>%
    summarise(is_um = mean(is_um), n = sum(n))%>%
    filter(n > 10)%>%
    ggplot(., aes(seg_dur_cat, is_um, color = sex))+
      geom_point()+
      geom_line(aes(group = sex))+
      ylim(0,1)+
      facet_wrap(~decade)

```

## Education

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width = "80%", fig.show = 'hold'}
  um_uh%>%
    mutate(DOB = year-age,
           edu = as.numeric(as.character(schooling)),
           higher = edu > 12,
           is_um = (word == "UM")*1)%>%
    filter(!is.na(edu), next_seg == "sp")%>%
    group_by(higher, sex, DOB, idstring)%>%
    summarise(is_um = mean(is_um))%>%
    ggplot(., aes(DOB, is_um, color = higher))+
      geom_point()+
      stat_smooth(se = F)+
      facet_wrap(~sex)

```